---
title: "Field Book Trait Reference Guide"
author: "Tuesday Crewsday"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
packages <- c("readxl", "dplyr", "knitr", "kableExtra")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed])
}

library(readxl)
library(dplyr)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This document provides a standardized reference for trait data collected using the Field Book application. It is intended to serve as supplemental documentation to support consistent data collection, trait interpretation, and downstream analysis.

This guide is designed to provide a centralized and accessible source for trait names, trait types, and trait definitions so that users can apply consistent terminology across projects.

The Shiny application was developed to support phenomic data handling and Field Book trait file generation through a graphical user interface (GUI). This application allows users to upload phenomic datasets and generate downloadable trait files formatted for use with Field Book.

# Purpose

The purpose of this reference guide is to:

- establish consistent terminology for all recorded traits
- provide a centralized reference for trait categories, names, and definitions
- support uniform data collection across users
- provide a low-barrier entry point for using Field Book and GRIN
- provide supplemental documentation for future reference and project use
- serve as a resource for GUI development

# Dataset Structure

Each entry in the dataset is defined by the following components:

- **Trait**: the trait name used for documentation and reference
- **Trait Description**: a written definition or explanation of the trait
- **Trait Category**: a grouping used to organize traits by type or application
- **Data Type**: the expected format of the recorded data

# Data Source

The data presented in this document are imported directly from the `maize_traits.xlsx` file and processed in R. Re-knitting this document after updates to this file will automatically refresh the reference table and maintain an up-to-date record of trait metadata.

```{r import-data}
# import data
maize_traits <- read_excel("maize_traits.xlsx")
```

```{r clean-data}
# clean up data
traits_clean <- maize_traits %>%
  rename(
    category = Catergory,
    data_type = `data type`
  ) %>%
  mutate(
    title = ifelse(is.na(title) | title == "", "Not Provided", title),
    description = ifelse(is.na(description) | description == "", "Definition not yet provided", description),
    category = ifelse(is.na(category) | category == "", "Not Specified", category),
    data_type = ifelse(is.na(data_type) | data_type == "", "Not Specified", data_type)
  ) %>%
  arrange(category, title)
```

# Trait Reference Table

The table below provides a reference list of Field Book traits for maize, including the trait, trait description, trait category, and data type.

```{r trait-table}
traits_clean %>%
  select(title, description, category, data_type) %>%
  kable(
    format = "html",
    caption = "Field Book Trait Reference Table",
    col.names = c("Trait", "Trait Description", "Trait Category", "Data Type"),
    escape = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

# Notes

This document is intended to function as a supplemental reference for Field Book trait documentation. It will be updated as the Field Book GUI, trait dictionary, and supporting metadata continue to be refined.