---
title: "Field Book Trait Reference Guide"
author: "Jazlyn Salazar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
packages <- c("readxl", "dplyr", "knitr", "kableExtra")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(packages[!installed])
}

library(readxl)
library(dplyr)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This document provides a standardized reference for trait data collected using the Field Book application. It is intended to serve as supplemental documentation to support consistent data collection, trait interpretation, and downstream analysis.

This guide is designed to provide a centralized and accessible source for trait names, trait types, and trait definitions so that users can apply consistent terminology across projects.

# Purpose

The purpose of this reference guide is to:

- establish consistent terminology for all recorded traits
- provide a centralized reference for trait categories, names, and definitions
- support uniform data collection across users
- provide a low-barrier to entry for using Field Book and GRIN
- provide supplemental documentation for future reference and project use
- serve as a resource for GUI development

# Dataset Structure

Each entry in the dataset is defined by the following components:

- **Trait Category**: a broad classification grouping traits by biological or experimental context
- **Trait Name**: the descriptor used within Field Book to record observations
- **Trait Type**: the data format used for recording values
- **Definition**: a brief explanation of what the trait represents and how it should be interpreted

# Data Source

The data presented in this document are imported directly from the trait categories and names excel file and processed in R. Re-knitting this document after updates to this file will (should) automatically refresh the reference table and maintain an up-to-date record of trait metadata.

```{r import-data}
# import data
traits <- read_excel("trait categories and names.xlsx")
```

```{r clean-data}
# clean and prepare data
traits_clean <- traits %>%
  rename(
    Trait_Category = Category,
    Trait_Name = Descriptor,
    Trait_Type = `Field Book Trait Type`
  )

if (!"Definition" %in% names(traits_clean)) {
  traits_clean <- traits_clean %>%
    mutate(Definition = "Definition not yet provided")
}

traits_clean <- traits_clean %>%
  mutate(
    Trait_Type = ifelse(is.na(Trait_Type) | Trait_Type == "", "Not Specified", Trait_Type),
    Definition = ifelse(is.na(Definition) | Definition == "", "Definition not yet provided", Definition)
  ) %>%
  arrange(Trait_Category, Trait_Name)
```

# Trait Reference Table

The table below provides a reference list of Field Book traits, including trait category, trait name, data type, and trait definition.

```{r trait-table}
traits_clean %>%
  select(Trait_Category, Trait_Name, Trait_Type, Definition) %>%
  kable(
    format = "html",
    caption = "Field Book Trait Reference Table",
    col.names = c("Trait Category", "Trait Name", "Trait Type", "Definition"),
    escape = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

# Notes

This document is intended to function as a supplemental reference for Field Book trait documentation. It will be updated as the Field Book GUI, trait dictionary, and supporting metadata continue to be refined.
 